---
title: "Assignment 1 Report"
output:
  pdf_document: default
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)

# Read the data
setwd("~/Library/CloudStorage/OneDrive-UvA/CLS/Experimental Design and Data Analysis/Assignment 1/EDDA_assignment1")
data <- read.table("crops.txt", header=TRUE)
data$County <- factor(data$County)  
data$Related <- factor(data$Related, levels = c("no", "yes"))
```

## Exercise 2

### a)

We first plot the data:

```{r, echo=FALSE}
ggplot(data, aes(x = County, y = Crops, fill = Related)) +
  geom_boxplot() +
  labs(title = "Crops by County and Related Factor",
       x = "County",
       y = "Crops") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

We then perform ANOVA only with County and Related:
```{r}
model = lm(Crops ~ County + Related, data=data); anova(model)
```

We follow by including the interaction between the two:
```{r}
model_int = lm(Crops ~ County * Related, data=data); anova(model_int)
```

The interaction terms are not significant, let's use the first model; are its residuals normally distributed?

```{r, echo=FALSE}
par(mfrow=c(1,2)); qqnorm(residuals(model))
plot(fitted(model), residuals(model))
```

Now we want to estimate the crops for a typical farm in County 3 for which landlord and tenant are not related. Let's look at the model results:
```{r}
summary(model)$coefficients
```

Thus, to calculate the desired value we compute:
$$Crops_{cou.=3, rel.=no} = 6800.6 + 959.7 = 7760.3$$
